### 2017-05-18 sections/implementation.tex
Update on Overleaf.

Let us assume that the matrix is not empty and is full rank. Then from the formula (\ref{eq:cr-def}) follows that if $r$ equals to the rank of matrix $\mtrx{A}$ the compression ratio is always higher than one. In other words the memory consumption of stored decomposition is bigger than the size of the original matrix. To make the compression algorithm applicable the parameter $r$ must conform to the condition

@@

\noindent
Considering the usual shape of matrix containing FEM results this inequality is easily satisfiable even for the $r$ being close to the rank of the original matrix as in the typical case the number of nodes or integration-points is much higher than the number of analysis steps and therefore $m<<n$.

@@

\paragraph{Approximation error}
In the usual case the most important parameter to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined approximation error threshold is not exceeded. To quantify the error the Mean squared error ($MSE$) metric is used. To be able to compare different data-sets having different scales the normalized version of $MSE$ is used - Normalized root-mean-square deviation ($NRMSD$). Both metrics are defined in section \ref{sec:error}.

To effectively calculate the final rank of the approximation matrix from the desired approximation error the interesting property of the singular values is used:

@@

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (\mtrx{A}_{ij})^{2} = \sum_{i=1}^{k}{s_{i}^{2}},
\label{eq:elem-sqr-sigma-sqr}
\end{equation}

\noindent
where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $\mtrx{M}$. The above formula states that the sum of squared elements of matrix $\mtrx{A}$ equals to the sum of squared singular values $s_{i}$ of the same matrix $\mtrx{A}$.

@@

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (\mtrx{A}_{ij} - \mtrx{A'}_{ij})^{2} = \sum_{i=r+1}^{k}{s_{i}^{2}},
\end{equation}

@@

Time complexity of the exact SVD decomposition algorithm is $O(m^2n)$, where $m<n$. This theoretical algorithm complexity is confirmed by two benchmarks where the dependency of execution time on varying matrix dimension is shown. The results of the benchmarks are depicted in Figure \ref{fig:ExeTime_rows} and Figure \ref{fig:ExeTime_columns}. Several observations were made from the results:

@@

Storage size of SVD decomposition itself can also be optimized. $\mtrx{S}$ being a diagonal matrix can be stored as single list of singular values $\sigma_{i}$ or can be even multiplied with the matrix of left singular vectors $\mtrx{U}$.

>>>>

Let us assume that the matrix is not empty and is full rank. Then from the formula (\ref{eq:cr-def}) follows that if $r$ equals to the rank of matrix $\mtrx{A}$ the compression ratio is always higher than one. In other words the memory consumption of stored decomposition is bigger than the size of the original matrix. To make the compression algorithm applicable the parameter $r$ must satisfy the condition

@@

\noindent
Considering the usual shape of matrix containing FEM results this inequality is easily satisfiable even for the $r$ being close to the rank of the original matrix as in the typical case the number of nodes or integration points is much higher than the number of analysis steps and therefore $m \ll n$.

@@

\paragraph{Approximation error}
In the usual case the most important measure to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined approximation error threshold is not exceeded. To quantify the error the Normalized root-mean-square deviation ($NRMSD$) is used. The normalized error metric enables to work with various data sets that have different scales. $NRMSD$ is defined in section \ref{sec:error}.

To effectively calculate the final rank of the approximation matrix from the desired approximation error the interesting property of singular values

@@

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (a_{ij})^{2} = \sum_{i=1}^{k}{s_{i}^{2}},
\label{eq:elem-sqr-sigma-sqr}
\end{equation}

\noindent
where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $\mtrx{A}$, is made use of. The above formula states that the sum of squared elements of matrix $\mtrx{A}$ equals to the sum of squared singular values $s_{i}$ of the same matrix $\mtrx{A}$.

@@

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (a_{ij} - a'_{ij})^{2} = \sum_{i=r+1}^{k}{s_{i}^{2}},
\end{equation}

@@

Computational complexity of the exact SVD decomposition algorithm is $O(m^2n)$, where $m<n$. This theoretical algorithm complexity is confirmed by two benchmarks where the dependency of execution time on varying matrix dimension is shown. The results of the benchmarks are depicted in Figure \ref{fig:ExeTime_rows} and Figure \ref{fig:ExeTime_columns}. Several observations were made from the results:

@@

Storage size of SVD decomposition itself can also be optimized. $\mtrx{S}$ being a diagonal matrix can be stored as single list of singular values $s_{i}$ or can be even multiplied with the matrix of left singular vectors $\mtrx{U}$.

### 2017-05-18 sections/introduction.tex
Update on Overleaf.

The effort to reduce size of resulting data from complex finite element analyses to accelerate (or even make possible) post-processing using common personal computer is not new. In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by continuous functions. These approximation functions can be then described by a small number of parameters. The focus is on geometry. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by a simple continuous function, e.g. linear function. Although this approach has great results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it cannot be determined in advance.

In this study it was decided to apply purely algebraic approach. Various methods were considered, e.g. discrete wavelet transform \cite{Lui2001} and discrete cosine transform \cite{Watson1994}, for their successful use in image compression. The reason why the SVD method was finally chosen is that it is well suited for all kinds of data and the implementation is straightforward, considering that the results from the FEM analyses can be viewed as a series of arbitrary rectangular matrices.

>>>>

Amount of data produced by a complex finite element analysis can be enormous and typical personal computer is not capable to store, process and visualize the results in reasonable time. Singular Value Decomposition (SVD) is a well known factorization method that provides rich information about matrix systems. One of its many applications is image compression where it can significantly reduce size of data representing image while preserving quality of image appearance.

The effort to reduce size of resulting data from complex finite element analyses to accelerate (or even make possible) post-processing using common personal computer is not new. In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by continuous functions. These approximation functions can be then described by a small number of parameters. The focus is on geometry. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by a simple continuous function, e.g. linear function. Although this approach has remarkable results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it cannot be determined in advance.

In this study it was decided to apply purely algebraic approach. Various methods were considered, e.g. discrete wavelet transform \cite{Lui2001} and discrete cosine transform \cite{Watson1994}, for their successful use in image compression. The reason why the SVD method was finally chosen is that it is well suited for all kinds of data and the implementation is straightforward, considering that the results from the FEM analyses can be viewed as a series of arbitrary rectangular matrices. Efficient data storage is connected with data structure \cite{Ivanyi2012, Ivanyi2014} and this area will be subject of further research.

The quality of any compression method depends on the nature of input data. Therefore, several different non-trivial finite element analyses are used as benchmarks for an implementation of the compression algorithm. The ability to control the quality of compression results is the key requirement for an implementation.

### 2017-05-18 sections/math.tex
Update on Overleaf.

The objective of a compression algorithm is to reduce amount of data representing FEM results and also the ability to reconstruct original data from its smaller representation. This saves storage capacity and also accelerates data transfer between computers as the analysis itself and the post-processing of results is usually done on different work stations.

Exact SVD of a $m \times n$ matrix has time complexity $O(min(mn^2, m^2n))$ using the "big-O" notation. In \cite{Holmes2007, Candes2011, Woolfe2008, Martinsson2011} are described randomized algorithms for constructing approximate matrix factorizations that have better asymptotic complexity. Their main purpose is to accelerate calculations of matrix decompositions which tend to be very slow for large data sets.

Compression method can be lossy or lossless based on quality of data reconstructed from its compressed representation. Lossless methods are able to fully reconstruct original data. Lossy methods on the other hand produce only approximations of original data.

@@

\begin{equation}
\mtrx{A}=\sum_{i=1}^{n} \mtrx{U_{i}}S_{i}\mtrx{V_{i}^{\mathsf{T}}},
\label{eq:svd-expansion}
\end{equation}

\noindent
where $S_{i}$ is the $i$-th singular value, $\mtrx{U_{i}}$ and $\mtrx{V_{i}}$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $S_{1} \geq S_{2} \geq S_{3} \geq ... \geq S_{n}$, the above formula implies that the first term of the sum would have the highest contribution and the last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

\begin{equation}
\mtrx{A'}=\sum_{i=1}^{r} \mtrx{U_{i}}S_{i}\mtrx{V_{i}^{\mathsf{T}}}.
\label{eq:svd-approx-expansion}
\end{equation}

The quality of approximation depends on the magnitude of the singular values omitted from the approximation formula, namely $S_{r+1} ...  S_{n}$. The compression algorithm is based on an assumption that the first singular value is order-of-magnitude higher than singular values at the end of the decomposition sequence. In special cases, when $r=n$ or $S_{i}=0$ for all $i \geq r$, the omitted singular values do not contribute to the sum and the compression is therefore lossless. In other cases, approximation error has to be calculated and taken into account to avoid loss of important details in data.

@@

\item \textbf{Rooted Mean Square Deviation}
\begin{equation}
RMSD=\sqrt{MSE}
\label{eq:rmsd-def}
\end{equation}

@@

PSNR is most commonly used to measure the quality of reconstruction of lossy compression methods (e.g. image compression). The signal in this case is the original data, and the noise is the error introduced by compression. PSNR is an approximation to human perception of reconstruction quality. This metric is not so important in area of FEM analyses where the human perception of visualizations is not as important as the exact mathematical accuracy of approximations. The reason to include PSNR in results is in particular to allow comparison with other image-related compression methods. PSNR is usually expressed in terms of the logarithmic decibel scale (dB).

\begin{equation}
PSNR=10\log_{10}\frac{(X_{max}-X_{min})^{2}}{MSE}
\end{equation}

\begin{equation}
PSNR=20\log_{10}\frac{X_{max}-X_{min}}{\sqrt{MSE}}=20\log_{10}\frac{1}{NRMSD}
\end{equation}

\begin{equation}
PSNR=-20\log_{10}NRMSD
\label{eq:psnr-def}
\end{equation}

>>>>

The objective of a compression algorithm is to reduce amount of data representing FEM results and also the ability to reconstruct original data from its smaller representation. This saves storage capacity and also accelerates data transfer between computers as the analysis itself and the post-processing of results is usually done on different workstations.

Exact SVD of a $m \times n$ matrix has computational complexity $O(min(mn^2, m^2n))$ using the "big-O" notation. In \cite{Holmes2007, Candes2011, Woolfe2008, Martinsson2011} are described randomized algorithms for constructing approximate matrix factorizations that have better asymptotic complexity. Their main purpose is to accelerate calculations of matrix decompositions which tend to be very slow for large data sets.

Compression method can be lossy or lossless based on quality of data reconstructed from its compressed representation. Lossless methods are able to fully reconstruct original data. Lossy methods, on the other hand, produce only approximations of original data.

@@

\begin{equation}
\mtrx{A}=\sum_{i=1}^{n} s_{i}\bf{u}_i\bf{v}_i^{\mathsf{T}},
\label{eq:svd-expansion}
\end{equation}

\noindent
where $s_i$ is the $i$-th singular value, $\bf{u}_i$ and $\bf{v}_i$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $s_{1} \geq s_{2} \geq s_{3} \geq ... \geq s_{n}$, the above formula implies that the first term of the sum would have the highest contribution and the last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

\begin{equation}
\mtrx{A'}=\sum_{i=1}^{r} s_{i}\bf{u}_i\bf{v}_i^{\mathsf{T}}.
\label{eq:svd-approx-expansion}
\end{equation}

The quality of approximation depends on the magnitude of the singular values omitted from the approximation formula, namely $s_{r+1} ...  s_{n}$. The compression algorithm is based on an assumption that the first singular value is order-of-magnitude higher than singular values at the end of the decomposition sequence. In special cases, when $r=n$ or $s_{i}=0$ for all $i \geq r$, the omitted singular values do not contribute to the sum and the compression is therefore lossless. In other cases, approximation error has to be calculated and taken into account to avoid loss of important details in data.

@@

\item \textbf{Rooted Mean Square Deviation}
\begin{equation}
RMSD=\sqrt{MSE}.
\label{eq:rmsd-def}
\end{equation}

@@

PSNR is most commonly used to measure the quality of reconstruction of lossy compression methods (e.g. image compression). The signal in this case is the original data, and the noise is the error introduced by compression. PSNR is an approximation to human perception of reconstruction quality. This metric is not so important in area of FEM analyses, where the human perception of visualizations is not as important as the exact mathematical accuracy of approximations. The reason to include PSNR in results is in particular to allow comparison with other image-related compression methods. PSNR is usually expressed in terms of the logarithmic decibel scale (dB)

\begin{eqnarray}
PSNR &=& 10\log_{10}\frac{(X_{max}-X_{min})^{2}}{MSE} =
\\
&=& 20\log_{10}\frac{X_{max}-X_{min}}{\sqrt{MSE}}=20\log_{10}\frac{1}{NRMSD} = \nonumber
\\
&=& -20\log_{10}NRMSD. \nonumber
\label{eq:psnr-def}
\end{eqnarray}

### 2017-05-18 sections/results.tex
Update on Overleaf.

Figures \ref{fig:ExeTime_rows} and \ref{fig:ExeTime_columns} show results from the benchmarks that are meant to test theoretical algorithmic complexity of the compression algorithm and its dependence on the size of input matrix. Outcomes are discussed in section Optimization.

>>>>

Figures \ref{fig:ExeTime_rows} and \ref{fig:ExeTime_columns} show results from the benchmarks that examine algorithmic complexity of the compression algorithm and its dependence on the size of the input matrix. Outcomes are discussed in section Optimization.
